---
title: "First_session"
author: "MartinGarlovsky"
date: "2021-11-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In the following classes we will be learning how to use `R` and `R studio` for data analysis. This will be a short introduction with the aim of getting you all familiar with the basics of how to use `R` for data management, exploring data, performing statistics, and plotting. There is a lot to learn in this short time but if you ever have any queries feel free to send me an [email](martin.garlovsky@tu-dresden.de) or find me in the Biology Building - room 258. 

### Outline for the course

* Downloading and installing `R` and `R studio`
* Basic arithmetic
* Objects
* Functions
* Vectors
* Data frames
* Scripts
* Loading data
* Packages
* The `tidyverse`

# What is `R`
Before we get started - what is `R` and why should you care?

`R` is a powerful, sophisticated and *free to use* piece of software to perform (almost) all your data analysis needs. From managing your data through to generating publication ready figures for that paper you've been working on. 

* `R` is perhaps the most widely used software used in Ecology and Evolution and has several key features:
  + Reproducible and stable
  + Free to use and open source
  + 

# Downloading and installing `R`

# First R session
Now we've installed `R` on to our computers lets open it up. You should see something like this. 

Not much to see here. But you're greeted with the:

`>` called the prompt.

### Things to know
* `R` is *case sensitive* - which means `R` know the difference between 'martin' and 'Martin' so that you'll get an error if you type `martin` when you mean `Martin`!
* Speaking of error messages - the error messages are often not very helpful for beginners. But the internet is your friend and a quick search will often help you troubleshooting.

## Simple arithmetic
We'll start with some simple arithmetic to get you familiar with how R works. You can type directly in to the `console` window to perform tasks. This lets you directly interact with R to do things - like perform calculations. 

* The math symbols:
  + `+` addition
  + `-` subtraction
  + `*` multiplication
  + `/` division
  + `^` raise to the power of

Type 1 + 1 and hit `enter`
```{r}
1 + 1
```

Now try 3 x 9  

```{r}
3*9
```

Try the following:

* 5 x 9
* $3^6$
* Add 3 to 11 and then multiply by 7
* Subtract 14 from from 110 and then raise to the power of 3 plus 1.4 (the answer should be `r (110 - 14)^(3 + 1.4)`)
* Divide 9 by 2.4
* $\sqrt51$
* $\log_{10}41$ (answer = `r log10(41)`)

```{r}
5 * 9
3 ^ 6
(3 + 11) * 7
(110 - 14) ^ (3 + 1.4)
9 / 2.4
sqrt(51)
log10(41)
```

## Objects
R is 'object orientated'... we use the `<-` assign symbol. You can think of this symbol as meaning "make the thing on the left (the object) in to the output of the command on the right. Objects store information, this can be things like sums or the output of a function, data frames or other functions themselves. 

Lets create an object called 'x' equal to 5 and hit enter.
```{r}
x <- 5
```

What happened? Nothing right? And you're left with the prompt blinking again... This is often a good sign in R! The object `x` has been stored in R's 'brain'. You can think of how making objects as "make `x` equal to 5".

* Note that `->` is equally valid but used less often. 
* A handy shortcut - instead of typing out `<` then `-` each time we can type '`alt` + `-`'.

You can inspect the object by typing `x` in the console and hitting `enter` again:

```{r}
x
```


We can perform operations on objects such as mathematics:
```{r}
x * 5
```

Note that while we performed an operation on `x`, it is itself unchanged. We can check this by inspecting `x` which is still 5 (not 25)
```{r}
x
```

We can also store the output as a new object
```{r}
answer <- x * 5
```

Be careful with `<-`; objects can be overwritten. So if you now assign something else to `answer` you will lose the previous information:

```{r}
answer <- 200
answer
```


These objects are stored in R's "brain". This means that when we type `x` or `answer` R will search for an object. We can see what objects are stored using `ls()` (short for list). 

```{r}
ls()
```

It is often good practice to clear R's brain at the beginning of every new session. This is because we may have 2 completely different projects where we have stored things with similar names (e.g. `model1`). We need to start fresh so there is no confusion. 

We can remove individual objects using `rm()` (short for 'remove'), or we can clear everything from R's brain by typing `rm(list = ls())` (don't worry too much about what this means for now).

```{r}
rm(x)

rm(list = ls())
```

Now see what information is stored in R's brain again, it should be blank:
```{r}
ls()
```


## Functions
Functions carry out procedures on numbers or objects. For instance, you've already used some functions when you calculated $\sqrt51$ and $\log_{10}41$. Functions take 'arguments' telling it what to use and what to do. For instance, we can use the `round()` function to round numbers, which takes as its arguments (i) a number and (ii) the number of decimal places to be used.

```{r}
round(x = 3.1415926535897932384626433832795028841971693993751058209749445923078164062862089986280348253421170679, digits = 3)
```

We can create objects which contain the output from functions

```{r}
log_value <- log(5299)
```

## Vectors

Vectors are the basic unit in R. A vector stores a 1-dimensional set of values (for instance the numbers 1 to 10). Vectors should contain only one type of data. That means all numbers, or all character strings, etc. 

### Numeric vectors
We can create a vector using the `concatenate` function which is shortened to simply `c()`.
```{r}
num_vec1 <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

Each entry in a vector is called an 'element'. We can perform operations on vectors. For instance we could add 1 to each element in `num_vec1`

```{r}
num_vec1 + 1
```

We could also have created the same vector using the `seq()` function. Notice that `seq()` takes multiple arguments (i.e. `from = `, `to = `, `by = `, etc.) and we can generate the same result in different ways. You can check what arguments can be used with a function with the `?` to access the help page for a function (i.e. `?seq`). Note that these 'help' pages are often not very useful for beginners, but as you learn more about R you will find it easier to interpret these pages.

```{r}
num_vec1a <- seq(from = 1, to = 10, by = 1)
num_vec1b <- seq(from = 1, to = 10, length = 10)
```

We can check whether our two different uses of `seq()` produce the same result by subtracting every value from `num_vec1b` from `num_vec1a`. We should get all zeros:  $1 - 1 = 0$, $2 - 2 = 0$, etc.). 

```{r}
num_vec1b - num_vec1a
```

This demonstrates again how operations performed on vectors work. The first element from `num_vec1b` (1) was subtracted from the first element of `num_vec1a` (1), and so on. 

### Character vectors
Character vectors work in a similar way to numeric vectors that we just introduced. But first a little terminology. `Martin` is a character 'string' containing a series of characters (letters) next to each other. 

We can create a character vector like so

```{r}
myname <- c('Martin', 'David', 'Garlovsky')
```

### Working with vectors
What if we want to access information from within our vectors? Here we'll also explain what the numbers that appear at the left hand side of the console, i.e. `[1]` mean. This number 1 in square brackets is telling us the position of this element in the vector. If we have a longer vector that prints over multiple lines we'll see more numbers telling us the position of that element in the vector. For instance below, the 19th element in our vector is the number `37`, while the 91st element is the number `181`. 

```{r}
seq(from = 1, to = 200, by = 2)
```

If we want to know the value from a specific position in our vector we use the square brackets (`[]`). For instance, if we want to find out what the 7th number in our numeric vector is:

```{r}
num_vec1[7]
```

Or if we wanted to find my middle name:

```{r}
myname[2]
```

You can also create vectors using the `:` symbol as a shortcut to get a vector of integer values. For instance, if we want the first 10 letters of the alphabet, we can combine the `letters` with the index `1:10`:

```{r}
letters[1:10]
```


## Data frames

R is really great for working with what are called 'data frames'. Data frames contain rows and columns of vectors containing information. Usually (hopefully) each row in a data frame corresponds to an observation (e.g. from a single individual or sample), and each column contains a variable (e.g. treatment, weight, Date, etc.). We'll make this clear with a dummy example to begin with. Let's imagine we have measured the height and weight of 6 people that took part in a study where volunteers ate a vegetarian or vegan diet for 6 weeks.

First, lets create the 3 variables; `height` (in cm), `weight` (in kg), and `diet` (vegan or vegetarian) by making vectors. 
```{r}
height <- c(165, 175, 162, 160, 189, 200)
weight <- c(90, 68, 110, 56, 89, 120)
diet <- rep(c('vegan', 'vegetarian'), 3)
```

Then we'll create the data frame using the `data.frame()` function.
```{r}
diet_data <- data.frame(height, weight, diet)
```

### Inspecting the data
Often the very first thing to do after loading our data is to make a quick first inspection. There are several useful functions for inspecting the structure and contents of data frames. Base R has many useful functions, and the `tidyverse` offers a range of alternatives (which we'll see later).

#### `str`
First lets take a look at the structure of the data using the `str()` function. By 'structure' we mean how many rows and columns are there in the data frame? What are the names of the variables in the data? What form are the variables (i.e., numeric, character, integer, etc.)?

```{r}
str(diet_data)
```

We can see that the data frame consists of 6 observations (`obs.`) of three variables. `height` and `weight` are numeric (`num`) variables and the `diet` variable consists of character strings (`chr`).

#### `head`
Along with checking the structure of the data, it is often useful to take a quick look at the data frame (rather than trying to see all the data at once). We can use the `head()` function to look at the first few rows of the data frame. By default `head()` returns the first 6 rows of the data frame. What about if we wanted to see more (or less) than 6 rows? - Hint: take a look at `?head`

```{r}
head(diet_data)
```

We see the first 6 rows of the data frame and all the variables. For larger data frames with more variables that extend beyond the console window the information may be printed over several lines. 

* Like `head()` we can also look at the last entries in a data frame using `tail()`.

#### `summary`

The `summary()` function can also be used to... summarise the data in a data frame. For instance, showing us the range of values and simple statistics like the minimum and maximum values, mean, median, and interquartile range. 

```{r}
summary(diet_data)
```

`summary()` is a versatile function and can be used on a range of different types of objects which we will learn about later. 


# Using the script window
So far we've been working exclusively in the `console`. Unlike in the console window, hitting `enter` in the script window no longer executes your code, but instead will insert a new line. To execute a line of code we have to highlight the text and press `cmd + enter` (on Mac) or `ctrl + enter` (PC). **To execute an entire line we do not have to highlight anything** but can simply hit `cmd + enter`/`ctrl + enter` and R will execute the code and move to the next line. In the next section introducing `packages` this will hopefully become clear.

## Annotating your script
Working with scripts has many advantages over working directly in the console. First, when working directly in the `consol`, if you make a mistake you have to go back and retype everything in to the console (or at least hit the `up` arrow and find your mistake again). Unless you save the `workspace` (don't), you will also not have a record of your workflow - which is crucial to make sure your analyses are ***reproducible***. Working with scripts also allows you to add annotation to your code to help future scientists (this may be your future self) know exactly what you did. The importance of annotating your scripts cannot be overstated! Writing your code in scripts - combined with adding detailed annotation - also allows you to share your code and work collaboratively. 

To add a comment to your script first type the `#` symbol followed by your annotation. You can also use the `#` to 'comment' out a line of code so that `R` knows not to try and evaluate this line of text. 

```{r}
# Here is an example of adding some annotations to a script. 

# A plot of the diet data using base R
plot(diet_data$height, diet_data$weight,
     col = c('red', 'blue'),
     pch = 16, 
     cex = 5, # increase the size of the points with 'cex'
     ylim = c(50, 130), # increase the Y axis limits so points are not cut off
     main = 'Height (cm) vs. Weight (kg)',
     xlab = 'Height (cm)', ylab = 'Weight (kg)',
     las = 2 # we can change the orientation of the axis labels which is a bit odd in this case...
     )

# We've not covered any statistics yet so this code is commented out
#cor.test(diet_data$height, diet_data$weight)

```


# Setting the working directory and importing data
So far we've used some 'dummy' data to demonstrate how we can use `R`. When it comes to using your own data, the first step is to load the data in to `R`. To understand how to do this we first need to understand a little about how the file system works on your computer and working with 'directories'. 

When we want load data in to `R` from our computer we first have to tell `R` where to 'look' for the file containing the data. By default, `R` will look in what is called the current *working directory*. A directory is computer language for a folder. A directory can contain files (like your data), or other folders (other directories). 

When we ask `R` to load the data it first looks in the current working directory. You can think of this like if you were to look for a pair of shoes to put on. The room you are currently in is the current working directory. The shoes (file) you want may be in this room (directory), but might be in another room (directory). If the folder (shoes) are in a different directory (room) to the one we are currently in we have to tell `R` to go and look in another directory.

We can find the current working directory with `getwd()` (`wd` stands for working directory)
```{r}
getwd()
```

This character string, separated by `/`'s is called a **`path`**, and gives the directions to the current working directory. So we are in the directory `Introduction2Rstudio` which is in the `Dropbox` directory, which is in the `martin` directory, which is in the `Users` directory, which is in the `root` directory.


You can see the current working directory and what files are in there interactively using the `Files` tab in the `help` panel. We could move the files we want to load in to `R` in to the current working directory. Alternatively, we can tell `R` to move to a different working directory to look for files using `setwd()`. Here I set the working directory to `"/Users/martin/Dropbox/TUDAZ"`, using the shortcut `~/` which tells `R` to start from my 'home' directory (`"/Users/martin"`).

```{r, eval = FALSE}
setwd('~/Dropbox/TUDAZ')
```

Then we can check that we have changed working directory with `getwd()` again
```{r, eval = FALSE}
getwd()
```

```{r, echo = FALSE}
print("/Users/martin/Dropbox/TUDAZ")
```

Alternatively, you can navigate anywhere in your computer using the file path, though this is a bit advanced for beginners and those not familiar with the path system.

## Loading the data
Once we have changed the working directory to the tell `R` where we can find our data, we can finally load the data in to `R`. There are many different file types, and you might be most familiar with something like an excel spreadsheet. Excel may be good for entering your raw data and there are packages available to load data directly from `.xls` files. However, often we will work with `.csv` or `.txt` files which have advantages over excel files such as taking up less space on your computer, making them faster to loadrs

We have a data set of bumble bee visits to different flowers in the form of 'comma separated values', or `.csv` file in the `data` folder. To load the data we can use the `read.csv()` function and use the `<-` symbol to call the loaded data `bumblebees`. 

```{r, echo = FALSE}
bumblebees <- read.csv('data/bumblebee_data.csv')
```

```{r, eval=FALSE}
bumblebees <- read.csv('bumblebee_data.csv')
```

For this example the only argument we need is the path to the `.csv` file.  If we try and load the same data saved as a 'tab-delimited text' `.txt` file using the `read.delim()` function we need to specify a couple more arguments. `sep = '\t'` tells the function that values in each column are separated by a 'tab' which in computer code is denoted with `\t`. We also need to tell `read.delim()` that the first row in the file should be used as column names using `header = TRUE`. We'll call this object `bumble_txt` and then quickly inspect the data.

```{r, echo = FALSE}
bumble_txt <- read.delim('data/bumblebee_data.txt', sep = '\t', header = TRUE)
```

```{r, eval=FALSE}
bumble_txt <- read.delim('bumblebee_data.txt', sep = '\t', header = TRUE)
```

```{r}
head(bumble_txt)
```


## Managing your projects and folders

Ideally, we want to organise our projects and data efficiently. It is a good idea to have separate folders for each project you are working on. For instance, you might have one folder for this `R` course, and another folder for your dissertation project. These should be in a `parent` directory called `TUDAZmasters` or something like that. `R studio` has some dedicated tools to make working in separate projects easier that we will introduce later. For instance, this document you are reading is part of an `R project`. All the data and R scripts relating to this course can be found in one directory (called `Introduction2Rstudio`).


# Packages
Next we'll introduce one of the things which makes `R` such a versitile and powerful tool for data analysis. That is the huge numbers of [`packages`](https://cran.r-project.org/web/packages/available_packages_by_name.html) which extend the utility of `R` for almost any data analysis or graphics problem you might have. 

## Installing and loading packages
When you first downloaded and installed `R` it came ready to use with some packages already installed (such as `stats` and `graphics`). But to really harness the power of `R` we can install additional packages. We will be using the `tidyverse` package and environment a lot in the following sections which we'll install here to demonstrate how packages work. First we use `install.packages()` to download the `tidyverse` package from the web. Then we'll use `library()` to load the package.

```{r, eval=FALSE}
install.packages('tidyverse')

library(tidyverse)
```

```{r, echo = FALSE}
library(tidyverse)
```

New users often have some difficulty understanding the way we work with packages. A common mistake is to use `install.packages()` each time we want to load a package, instead of `library()`. You can think of how we use `R` combined with `packages` like installing and using software on your phone or computer. The first time you wanted to listen to music on your phone you downloaded and installed the Music app. That is like using `install.packages()`. Every time you want to listen to music you do not download the app again, you just open the app by pressing the icon. That is like using `library()`. You need to be connected to the internet to use `install.packages()`, whereas you can be completely offline when you use `library()` (and most of the time when using `R`). 


# The `tidyverse`
The [`tidyverse`](link) is a group of packages designed to work together to make data management and analysis easier. There are many other ways to write your code and no one way is superior to the other. The `tidyverse` follows a set of principle which you may find agreeable, or somewhat dogmatic. 

## tidy data
One of the key tenets of the tidyverse is 'tidy' data. You may remember that earlier we mentioned that ideally data should be in the form of having each row corresponding to an 'observation' and each column a 'variable'. This is the tidy data structure. It may not always be feasible or indeed desirable to record your data by hand in the tidy format. However, once you load your data in to `R` we can easily 'reshape' the data appropriately. 

To demonstrate, consider the following data frame which records the weight of our volunteers before and after the 6 week vegan or vegetarian diet treatment. 

```{r, echo=FALSE}
diet_2 <- data.frame(diet = rep(c('vegan', 'vegetarian'), each = 3),
                     weight_1 = rnorm(6, mean = 70, sd = 2),
                     weight_2 = rnorm(6, mean = 70, sd = 1)) %>% 
  mutate(across(2:3, ~ round(.x, 1)))

diet_2
```

This data is not in the tidy format as the `weight` variable is split over two columns, `weight_1` and `weight_2`, corresponding to the two time points measurements were taken of each individual before and after the experiment. By comparison a tidy version of the same data would look like the following, with one observation per row, and a new variable (column) indicating the measurement `time`. 

```{r, echo=FALSE}
diet_2 %>% 
  pivot_longer(cols = 2:3, names_to = 'time', names_prefix = 'weight_', values_to = 'weight') %>% 
  as.data.frame()
```

## Penguin data set
There are vast amounts of data available online for you to download and explore. For our short intro we'll use a data set containing measurement of penguins on the Palmer Archipelago which is a nice example data set to be able to demonstrate all the various things we need to be able to do. This data set has superseded the previously popular `iris` data set collected by R. A. Fisher due to his (and his contemporaries) association with [eugenics](https://nautil.us/issue/92/frontiers/how-eugenics-shaped-statistics). 

First we have to download the data which is available as part of the `palmerpenguins` package, collected and made available by Dr. Kristen Gorman and the Palmer Station. Then we'll load the data and take a look. Notice that I have 'commented out' the first line of code where we install the package - as I already installed the `palmerpenguins` package on this computer. 

```{r}
#install.packages('palmerpenguins')

library(palmerpenguins)

str(penguins)
head(penguins)
summary(penguins)
```

We can see the `penguins` data is in the tidy format (yay) and consists of 344 observations of 8 variables describing penguins by:

* species of the penguin 
* island it was found on
* bill and flipper measurement
* body mass, sex, and collection year

Using the `summary()` function we can also see that we have some `NA` values. `NA` stands for 'not available' and is `R` notation for missing values. Missing values are important as they can affect how functions work. It is also important to distinguish missing values (`NA`s) from 'real' `0`s. 

## `dplyr`

`dplyr` uses a set of 'verbs' to manipulate your data frames. The main players are:

* `select` - to choose variables
* `mutate` - to add new variables
* `filter` - to subset the data
* there are many other very useful `dplyr` verbs

All `dplyr` functions take as their first argument the data frame to manipulate, followed by one or more other arguments. 

### `select`
We can use the `select()` function to include or exclude certain variables (columns) from the data frame. For instance, if we are only interested in data to compare bill measurements between species on each island. The first argument in the `select()` function is the data frame to be manipulated, and the following arguments are the names of the variables we wish to select. 

```{r}
select(penguins, species, bill_length_mm, bill_depth_mm)
```

`select()` also allows us to rename variables at the same time as selecting them. For instance, we might want to make the names shorter so we have less to type later. At the same time we'll assign this subset of the data to a new data frame called `bill_data`. Note that using `select()` to include only some variables does not affect the original `penguins` data. 

```{r}
bill_data <- select(penguins, species, length = bill_length_mm, depth = bill_depth_mm)
```

Lets look at the new data frame:

```{r}
head(bill_data)
```

And make sure the original data is unchanged using the `dim()` function (short for dimensions), which should show us that we still have 344 observations of 8 variables:

```{r}
dim(penguins)
```

`select()` can be used in many different ways that we do not have time for now. But briefly, note that we could have gotten the same resulting data frame by specifying which variables we **didn't** want to include using `-` before the variable name:

```{r}
select(penguins, -island, -flipper_length_mm, -body_mass_g, -sex, -year)
```

### `mutate`
We will often want to perform some sort of transformation on our data. For instance, converting a variable in to different units, or performing a log transformation. As with `select()`, `mutate()` takes the data frame we want to manipulate as the first argument, followed by an expression containing the instructions for what we want to achieve. Let's use `mutate()` to convert `body_mass_g` in to kilograms. 

```{r}
mutate(penguins, body_mass_g/1000)
```

We now have a new column ``body_mass_g/1000`` in the data frame. However, as we did not assign the output of `mutate()` to an object, we cannot access these results. ``body_mass_g/1000`` is also a messy variable name. We can designate a name for the new variable in `mutate()` with the `=` symbol and assign the output of `mutate()` to a new object. `mutate()` can be used to create more than one new variable at a time, by separating each new variable by a comma. To make the code easier to read we can also write each on a new line.

```{r}
penguins_kg <- mutate(penguins, 
                      body_mass_kg = body_mass_g/1000,
                      bill_area = bill_length_mm * bill_depth_mm)
```

### `filter`
We can use `filter()` to subset the data to include only rows which meet conditions that we choose. To specify conditions we need to become familiar with using relational and logical operators.

* Relational operators are used to compare between variables such as:
  * `x < y`: is x less than y
  * `x <= y`: is x less than or equal to y
  * `x > y`: is x more than y
  * `x >= y`: is x more than or equal to y
  * `x == y`: is x equal to y
  * `x != y`: is x not equal to y

Be careful to note that `==` is not the same as `=`. In `R` the single equals (`=`) is often used within functions, whereas the double equals (`==`) is used to test for equivalence (i.e. how you might normally think of `=` symbol).

* Logical operators 
  * `x | y`: is x OR y true?
  * `x & y`: are x AND y true?

Perhaps we are only interested in female penguins? Again, `filter()` takes the data frame we are to manipulate as its first argument, followed by the conditions which must be met for inclusion. 

```{r}
filter(penguins, sex == 'female')
```

Or... maybe we are interested in all penguins with a flipper length greater than 200mm?

```{r}
filter(penguins, flipper_length_mm > 200)
```

Finally, we can combine conditions. For instance, if we want to subset the data for male Adelie penguins we separate each condition with a comma. 

```{r}
filter(penguins, species == 'Adelie', sex == 'male')
```

We can also specify logical conditions to combine criteria. Let's say we want to only look at penguins with really big or really small bills. We use the `|` (OR) operator in the call to `filter()`. Remember that if we want to save the new 'filtered' data frame we have to assign the results to a new object.

```{r}
extreme_bills <- filter(penguins_kg, bill_area > 800 | bill_area < 660)
```


### `%>%` or `|>`: the pipe
What if we want to `select()` only a few of the variables, `mutate()` the data and `filter()`? The 'pipe' symbol `%>%` or `|>` allows us to link multiple functions together in a few lines of code in a way that is easily readable. 

We can think of how `%>%` works as "take the thing on the left hand side (i.e. a data frame), and perform the thing on the right. 

For instance, we can filter our `penguins` data frame by putting it through the `%>%`. Note that we no longer have to specify the data frame as the first argument in `filter()`:

```{r}
penguins %>% filter(species == 'Adelie', sex == 'male')
```

Now lets combine `dplyr` functions to first `mutate()` the data to create a new variable called `scaled_flipper`, and then `filter()` the data based on this new variable. 

```{r}
penguins %>% 
  mutate(scaled_flipper = flipper_length_mm/body_mass_g) %>% 
  filter(scaled_flipper > 0.05)
```

You can see that we can put each function on a new line. `R` will take the `penguins` data, and first perform our `mutate()` function, and afterwards perform the `filter()`. If we were to try and perform the `filter()` first we would get an error message as the new new variable created by `mutate()` doesn't exist yet:

```{r, error=TRUE}
penguins %>% 
  filter(scaled_flipper > 0.05) %>% 
  mutate(scaled_flipper = flipper_length_mm/body_mass_g)
```


### `group_by` and `summarise`
The functions `group_by()` and `summarise()` are often useful in combination and will help us demonstrate some `dplyr` functionality. By combining these two functions we can get manipulate our data frame to get some descriptive statistics about our penguin data. For instance, we might want to know the mean weight of penguins from each species. As with all the `dplyr` functions, both `group_by()` and `summarise()` need the data frame as the first argument unless we are using pipes. 

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(mn_weight = mean(body_mass_g))
```

Why are the values for `Adelie` and `Gentoo` '`NA`'? We've come across a problem with how the `mean()` function is treating missing values (`NA`'s) in the data. 

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(mn_weight = mean(body_mass_g, na.rm = TRUE))
```

## `ggplot2`

Now we've learned how to laod our data in to `R`, inspect the data, and manipulate the data frame, it's time to start making some graphs! `R` is **very** good for making high quality figures. As with many things in `R` there are a million and one ways to approach the same problem and it is often a matter of personal preference which way to do things. You can make plots in `Base R` but here we'll introduce using `ggplot2`, which is another part of the `tidyverse`. 

The `gg` in `ggplot2` stands for '**g**rammar of **g**raphics'. When using `ggplot2`, we have to specify: 

* `data`
* `aes` (short for 'aesthetics')
* `geom` (short for geometry)

`ggplot2` works by adding the various layers of your plot on top of each other. We do this by first calling the `ggplot` function, which like all `tidyverse` functions, takes the `data` as its first argument. Then we set the `aes`thetics we want to visualise, and then add `geom` layers with the `+` symbol. To illustrate what all this means lets look at a scatterplot of the penguin data.

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point()
```
```{r, echo=FALSE}
#ggsave(filename = 'output/first_ggplot.pdf')
```

First we use the `ggplot` function and specify the data (`penguins`) to be used. Then we tell `ggplot` what aesthetics (`aes`) we want. We want `body_mass_g` on the x-axis, `bill_length_mm` on the y-axis, and points coloured according to `species.` Finally, we want to draw a scatterplot so we specify the `geom` as `geom_point`.

Not bad... but this is hardly ready for submitting to a journal for publication. The axis titles are ugly and the default colour choices are no good for certain kinds of colourblindess - and we want as many people to be able to interpret our figure as possible right? Maybe we don't like the grey background either.

We can start building on our original plot by adding more layers. We'll use `labs` to manually decide on the axis labels, and we can use `scale_colour_...` to change the colour of points. Here I have used `scale_colour_...` as there are multiple different `scale_colour_...` arguments. The last argument we've added is `theme_bw()` (where `bw` stands for black and white), which changes the overall 'theme' of the plotting window. There are many themes available, and you can additionally change almost any aspects of your plots manually. 

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)') + 
  scale_colour_viridis_d() +
  theme_bw()
```
```{r, echo=FALSE}
#ggsave(filename = 'output/second_ggplot.pdf')
```

To show just how much you can customise your graphics lets try an extreme example...
```{r, fig.width=10, fig.height=10}
ggplot(data = penguins, aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(colour = species)) +
  geom_smooth(aes(lty = island), colour = 'orange') +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)') + 
  scale_colour_viridis_d() +
  theme(legend.position = 'top',
        legend.title = element_text(face = 'bold', angle = 45),
        axis.title.x = element_text(colour = 'dodgerblue', angle = 45, size = 15),
        axis.title.y = element_text(colour = 'green', angle = 200, size = 15),
        axis.text = element_text(angle = 135, size = 15, colour = 'red'),
        panel.background = element_rect(fill = "lightblue",
                                        size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 1.5, linetype = 'solid',
                                        colour = "hotpink"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'dashed',
                                        colour = "purple")
        )
```
```{r, echo=FALSE}
#ggsave(filename = 'output/bad_ggplot.pdf')
```


# Plotting in `Base R`
One of the most powerful and appealing aspects of using R is the ability to make professional quality figures relatively easily, with huge flexibility. It is worth pointing out briefly some of the plotting tools available with your base installation of `R` before moving on to using `ggplot2`. You can essentially make a figure look however you want; you can even plot your data as [cats, smiley faces or rainbows...](https://github.com/Gibbsdavidl/CatterPlots)

## `hist`
`hist` is a really useful plotting tool to quickly plot histograms. This is particularly useful when you want to look at the distribution of values, for instance when checking that the assumption of normality holds for your linear models. To access a variable from a data frame we use the `$` after the data frame. 

```{r}
hist(diet_data$height)
```

Manually specifying the `breaks = ` argument can be important when plotting histograms as using the wrong 'bin' size can hide underlying structure in the data. 

```{r}
hist(diet_data$height, breaks = 25)
```

## `plot`
We can make a simple scatter plot with the `plot` function. As with many functions, `plot` can be used in different ways, take different arguments and will plot different things depending on the input. 

```{r}
plot(diet_data$height, diet_data$weight)
```

We can customise the look of plots almost endlessly. 

```{r}
plot(diet_data$height, diet_data$weight,
     col = c('red', 'green'),
     pch = 16,
     main = 'Height (cm) vs. Weight (kg)',
     xlab = 'Height (cm)', ylab = 'Weight (kg)')

```

## `boxplot`
Boxplots are good for comparing continuous variables between two or more groups. 
```{r}
boxplot(diet_data$weight ~ diet_data$diet,
     col = c('red', 'green'),
     main = 'Effect of diet treatment on weight',
     xlab = 'Diet treatment', ylab = 'Weight (kg)')
```
